---
title: "ReconST: gene-panel selection in one call"
output:
  html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{ReconST: gene-panel selection in one call}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

# Setup

ReconST learns gene weights on scRNA-seq reference expression and evaluates
reconstruction on spatial expression. Supply two Giotto objects and receive the
spatial object with gene scores attached. Spatial coordinates are not required
by the algorithm, and reference/spatial cells do not need to correspond.

Install once from the public ReconST repository:

```{r install}
library(Giotto)
installGiottoReconSTEnvironment()
```

At the start of each new R session, before creating Giotto objects or invoking
Python, select the environment:

```{r environment}
GiottoClass::set_giotto_python_path(python_path = "giotto_reconst_env")
```

Restart R if another Python environment has already been initialized. If ReconST
is already installed in your active Giotto Python environment, use
`envname = NULL` in `runReconST()` instead.

# Run

The example assumes you already have `gobject_sc` and `gobject_spatial`, each
with a `raw` expression matrix. Use matching, unique gene symbols or Ensembl IDs
in both objects. ReconST does not translate between identifier conventions.

```{r run}
gobject_spatial <- runReconST(
  gobject_sc = gobject_sc,
  gobject_spatial = gobject_spatial,
  output_dir = "reconst_results",
  seed = 42
)
```

This performs export, training, evaluation, and result annotation. Intermediate
input files are temporary. Results remain in `output_dir`; use a separate output
folder to preserve each run. The wrapper converts integer parameters for Python.
Training uses one CPU thread by default for compatibility with R's native
libraries; `num_threads` overrides this for the call and the previous Torch
thread setting is restored afterwards.

The default reference filters require 200 detected genes per cell and 100 cells
per gene. Reference counts are normalized to 10,000, while spatial expression
is evaluated as supplied. Confirm that these scales fit your data. On small or
already prepared reference data you can disable filters with
`min_genes_per_cell = NULL` and `min_cells_per_gene = NULL`, and disable reference
normalization with `normalize_target_sum = NULL` if appropriate.

Slots default to each object's active spatial unit and feature type. Use
`spat_unit` / `feat_type` and `spat_unit_sc` / `feat_type_sc` to select them
explicitly for the spatial and reference objects, respectively.

# Retrieve the panel

```{r panel}
meta <- fDataDT(gobject_spatial)
panel <- meta$feat_ID[which(meta$reconst_selected)]
writeLines(panel, "reconst_panel.txt")
```

Feature metadata contains `reconst_importance` and `reconst_selected`. Unscored
genes receive `NA`. Pass `subset_to_panel = TRUE` to return only selected genes.
The result CSV also preserves the original IDs in `gene_id` for reliable matching.

# Existing h5ad files

For a file-based workflow, the same function returns a summary list:

```{r files}
summary <- runReconST(
  input = list(sc_h5ad = "reference.h5ad", spatial_h5ad = "spatial.h5ad"),
  output_dir = "reconst_results"
)
gobject_spatial <- loadReconSTResults(gobject_spatial, summary)
```

# Small reproducible example

The following synthetic counts demonstrate the complete workflow without a data
download. They are only a software check, not a biological benchmark. Set up the
Python environment before running this block.

```{r synthetic-example}
set.seed(42)
genes <- paste0("gene", seq_len(20))
reference_counts <- matrix(rpois(20 * 40, 4), nrow = 20,
  dimnames = list(genes, paste0("ref", seq_len(40))))
spatial_counts <- matrix(rpois(15 * 20, 4), nrow = 15,
  dimnames = list(rev(genes[1:15]), paste0("sp", seq_len(20))))
gobject_sc <- createGiottoObject(expression = reference_counts)
gobject_spatial <- createGiottoObject(expression = spatial_counts)

result <- runReconST(
  gobject_sc = gobject_sc,
  gobject_spatial = gobject_spatial,
  output_dir = file.path(tempdir(), "reconst_example"),
  num_epochs = 2, embedding_size = 5, batch_size = 16,
  min_genes_per_cell = NULL, min_cells_per_gene = NULL,
  normalize_target_sum = NULL, device = "cpu"
)
fDataDT(result)
```

The installer defaults to a pinned, public ReconST revision. Local source
folders can be selected with `package_path`; use `reconst_version` to override
with a Git installation URL or a published PyPI version. Giotto provides the
file adapter in `inst/python/python_reconst.py`; the independent ReconST package
provides the model, training and gene-selection implementation.

A fixed seed improves repeatability within the same environment; hardware and
package versions can affect results. This tutorial requires user-supplied data
and is not executed during vignette rendering.
